---
title: "generalized mixed linear model: marginal versus conditional effects"
author: "Ward B. Eiling"
format: html
editor: visual
bibliography: GLMM_references.bib
---

## Generalized Mixed Linear Model with Non-linear Logit Function

In this document, we will reproduce the figure made by Dimitri Rizopoulos in de course "CE08", where the generalized mixed linear model (GLMM) with a non-linear logit function results in a discrepancy between the marginal and conditional relationships (see also @Zeger1988 for the similar and original treatment). Let us consider the following GLMM

$$
log \frac{\pi_{it}}{1-\pi_{it}} = \alpha_0 + \alpha_1 X_{it} + b_{0i} + \epsilon_{it} \quad b_{0i} \sim \mathcal{N}(0,\sigma_{b0}^2)
$$

Where $\pi_{it} = Pr(Y_{it} = 1)$ the probability of a positive response for subject $i$ at time $t$, $Y_{it}$ is the binary response variable, $X_{it}$ is the continuous covariate, $A_{it}$ is the binary treatment indicator, and $b_{0i}$ is the random intercept. The fixed effects are $\alpha_0$ and $\alpha_1$ for the intercept and the continuous covariate. The random intercept $b_{0i}$ is assumed to be normally distributed with mean zero and variance $\sigma_{b0}^2$.

We specify the following parameter values

-   $\alpha_0 = 0$
-   $\alpha_1 = 0.9$
-   $\sigma_{b0}^2 = 4$

For simplicity, this model does not include a treatment effect $A_{it}$ nor exogenous noise $\epsilon_{it}$. Now we will simulate data from this model and directly infer the relationships between the continuous covariate $X_{it}$ and the probability of a positive response $\pi_{it}$.

```{r}
#| label: fig-marginal-vs-conditional
#| fig.width: 10
#| fig.height: 8

set.seed(123) # For reproducibility

# Parameters
alpha_0 <- 0 # for simplicity
alpha_1 <- 0.9
sigma_b0 <- sqrt(4)

n_individuals <- 20 # Number of individuals to plot
X_grid <- seq(-5, 5, length.out = 10000) # Values for X_it

# Generate random intercepts for individuals
b0 <- rnorm(n_individuals, mean = 0, sd = sigma_b0)

# Conditional-level logit and probability: the mean individual (b0 = 0)
logit_conditional <- alpha_0 + alpha_1 * X_grid + 0
pi_conditional <- 1 / (1 + exp(-logit_conditional))

# Individual-specific logits and probabilities
logit_individuals <- sapply(b0, function(b) alpha_0 + alpha_1 * X_grid + b)
pi_individuals <- 1 / (1 + exp(-logit_individuals))

# Population-level logit and probability: average probabilities across individuals
pi_population <- rowMeans(pi_individuals)

# save the upcoming plot
png("figure/fig-marginal-vs-conditional.png", width = 1200, height = 1000, res = 150)

# Plot the conditional-level curve
plot(X_grid, pi_conditional, type = "l", lwd = 4, col = "black",
     ylim = c(0, 1), xlim = c(-5, 5), xlab = expression(X[it]), ylab = expression(pi),
     main = "Population, Conditional and Individual Logit Curves")

# Add individual curves (grey lines)
for (i in 1:n_individuals) {
  lines(X_grid, pi_individuals[, i], col = "grey", lwd = 1)
}

# add the population-level curve (red line)
lines(X_grid, pi_population, col = "red", lwd = 4)

# Add a legend
legend("topleft", legend = c("conditional", "population", "individuals"),
       col = c("black", "red",  "grey"), lty = c(1, 1, 1), lwd = c(4, 4, 1))

dev.off()
```

When we estimate the GLMM with the log link function, we obtain the conditional-on-the-random effects interpretation of fixed-effects regression coefficients (i.e., the effects of covariates on changes in an individual’s transformed mean response). This is directly opposed by marginal models such as GEE, where "population averages are the targets of inference" (Rizopoulos, CE08, p. 331)

## Mixed Linear Model: Generalized Linear Mixed Model with Identity Link Function

```{r}

```
