---
format: # docx
  pdf:
    papersize: a4
    fig-pos: 'H'
    tbl-pos: 'H'
    # keep-tex: true
    # toc: true
    number-sections: true
    colorlinks: true
    indent: true
    header-includes:
      - \usepackage{fancyhdr}
      - \usepackage{amsmath}
      - \usepackage{multirow}
      - \usepackage{booktabs}
      - \usepackage{pifont}
      # - \usepackage{float}
      # - \usepackage{tikz}
      # - \usepackage{subcaption}
    geometry:
      - top=2.5cm
      - bottom=2.5cm
      - left=2.5cm
      - right=2.5cm
    mainfont: Latin Modern Roman
    sansfont: Latin Modern Roman
    linestretch: 2
    fontsize: 11pt
    # no page numbers
link-citations: true
header-includes:
      - \pagestyle{empty}
csl: apa.csl
execute:
  message: false
  warning: false
fig-cap-location: top
---

```{r}
#| label: Packages
#| echo: false

# for knitting the document
library(papaja)
library(rlang)
library(rmarkdown)
library(knitr)
library(kableExtra)
library(here)
# for general data manipulation and plotting
library(tidyverse)
# for estimation
# library(lme4)
# library(geepack)
# for diagrams
library(dagitty)
library(ggdag)
library(DiagrammeR)
library(magick)
library(pdftools)
```

::: {#tbl-simulation-results}
```{r}
#| label: simulation-results
#| echo: false
#| cache: true
#| escape: false

results <- readRDS(here("simulation_results/GM123ad-1000reps-researchreport-cleanscript_nrep10000/results_beta0_bias_sd_success.rds"))

# remove gee_ex and gee_ar1 and gee_ind
results <- results[, !grepl("gee_ex|gee_ar1|gee_ind", colnames(results))]
# remove GM2
results <- results[results$GM != 2, ]
# rename GM = 3 to "G", GM = 1 to "A", GM = 3a to "B" and GM = 3d to "C"
results$GM <- ifelse(results$GM == 3, "G", 
                     ifelse(results$GM == 1, "A", 
                            ifelse(results$GM == "3a", "B", 
                                   ifelse(results$GM == "3d", "C", NA))))

# compute MC-SE
results <- results %>%
  mutate(mc_se = mlm_beta0_sd / sqrt(10000))

# reorder columns
results <- results[, c(1:5,7,6)]

# reset row numbers
rownames(results) <- NULL
# reorder so that model GM is in the order G, A, B, C

results <- results[c(7:12, 1:6, 13:24), ]

# Create a kableExtra table with a second header
kableExtra::kbl(results, "latex", booktabs = T,  escape = F, digits = c(0, 0, 0, 3, 3, 3, 3), col.names = c("GM", "T", "N", "Bias", "SD", "MC-SE", "SR"), row.names = FALSE) %>%
  add_header_above(c(" " = 3, '$\\\\beta_0$' = 3, " " = 1), escape = F) %>%
  kable_styling(full_width = TRUE) %>%
  column_spec(1, width = "5em") %>%
  collapse_rows(columns = 1:2, valign = "middle") %>%
  kableExtra::kable_classic_2()
```

\vspace{2em}

*Note.* GM: generative model. T: number of timepoints. N: sample size. Bias: $\frac{1}{n_{\text{sim}}} \sum\limits_{i = 1}^{n_{\text{sim}}} \hat{\beta}_{0i} - \beta_{0}$, which represents the difference between the mean of the estimated parameter values $\hat{\beta}_{0}$ and the prespecified treatment effect $\beta_{0} = 1$. SD: $\sqrt{\frac{1}{(n_{\text{sim}} - 1)} \sum\limits_{i=1}^{n_{\text{sim}}} \left(\hat{\beta}_{0i} - \bar{\beta}_{0}\right)^2}$, which is the standard deviation of estimates across replications. MC-SE: $\frac{\text{SD}}{\sqrt{n_{\text{sim}}}}$, which represents the Monte Carlo SE of bias. SR: model fitting success rate. 

Treatment effect bias for Generative Models G, A, B and C over 10,000 replications
:::
